---
title: "Next-generation sequencing - analysis with DESeq2"
subtitle: "Comparison of gene expression in DSCs after irradiation - functional analysis"
author: "Marc Bender"
date: '`r Sys.Date()`'
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../Results",
      output_format = "pdf_document"
    )
  })
output: 
  word_document:
    toc: true
    keep_md: yes
  pdf_document:
    toc: true
    fig_caption: yes
    keep_md: yes
  html_notebook:
    theme: united
    toc: true
  html_document:
    df_print: paged
    toc: true
    code_folding: "hide"
header-includes: 
- \usepackage[utf8]{inputenc} 
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage{float}
- \let\origfigure\figure
- \let\endorigfigure\endfigure
- \renewenvironment{figure}[1][2] {\expandafter\origfigure\expandafter[H]} {\endorigfigure}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  dev = "cairo_pdf"
)
```

```{r load_libraries, include = FALSE}
library(DESeq2)
library(tidyverse)
library(ekbSeq)
library(rrvgo)
library(pathfindR)
library(ggwordcloud)
```

```{r custom_functions, include = FALSE}
## wrapper around ggwordcloud function to allow for character strings as mappings
wordcloud_string <- function(data, nrow = 1, max_size = 30,...){
  args <- list(...)
  args <- lapply(args, function(x) if (rlang::is_string(x)) sym(x) else x)
  set.seed(42)
  ggplot(data, aes(!!!args)) +
    geom_text_wordcloud(area_corr = TRUE) +
    scale_size_area(max_size = max_size, trans = power_trans(1/.7)) +
    facet_wrap(~Direction, nrow = nrow) + 
    theme_minimal()
}
 
## ggwordcloud for pathway analyses
ggwordpath <- function(data, z, score, n, dec = FALSE,...){
  # data$Direction <- ifelse(data$ID %in% activatedGO, "Activated", "Repressed")
  data <- data[order(data[[score]], decreasing = dec), ]
  
  ## extract the top 30 up and downregulated pathways
  cloudInput <- rbind(
    head(data[data[[z]] == "Activated",],n),
    head(data[data[[z]] == "Repressed",],n)
  )
  wordcloud_string(data = cloudInput, label = "Term_Description", size = score,...)
}
```

```{r load_data}
setwd("..")
rdata <- "DSC_NGS_UV_adjusted.RData"
load(file = rdata)
```

\newpage 
# Resources and introduction

This document includes the functional analysis for the comparison of RNASeq data from dermal stem cells (DSCs) and Melanocytes. The exploratory data analysis and differential expression analysis are included in separate files. Statistical methods are described in a separate .docx file. 

For additional info on analysis workflows check: 
DESEQ workflow: https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html 
https://bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html  
https://www.bigbioinformatics.org/r-and-rnaseq-analysis 
https://github.com/bigbioinformatics/r-programming-and-rnaseq-workshop 
ClusterProfiler: https://pubmed.ncbi.nlm.nih.gov/34557778/ 

<!-- Sanity Check:  -->
<!-- ```{r Sanity_check} -->
<!-- sanity_check(rdata) -->
<!-- ``` -->

\newpage
# Pathway analysis (Wikipathways)

Pathway analysis was conducted with the R package pathfindR combining a classical overrepresentation analysis with the information gained from protein interaction networks to take into account semantic relationships between differentially expressed genes. In total `r nrow(allResPath)` genes were expressed above a threshold of 10 counts in at least 5 samples. Out of these `r nrow(allResPath) - nrow(pathfindR_input)` did not have a documented gene symbol, leaving ``r nrow(pathfindR_input)` genes as input for the pathfindR workflow. Out of `nrow(sigRes)` differentially expressed genes, `r nrow(pathfindR_input[pathfindR_input$padj < 0.05,])` genes were kept in the workflow. xxx (have to be manually curated from the pathfindR output) genes did not have any documented protein interaction in the PIN (protein interaction network), leaving xxx (have to be manually curated from the pathfindR output) genes as final input for the pathfindR enrichment. 

(For the publication: pathfindR messages with number of genes into results.)

```{r pathway_total_wikiPW, fig.width = 15, fig.height=24, fig.cap = "Pathway scores between irradiated and unirradiated DSCs. All pathways."}
scoreMatrix <- score_terms(
  enrichment_table = outputDf,
  exp_mat = expMat,
  cases = cases,
  use_description = FALSE, # default FALSE
  label_samples = FALSE, # default = TRUE
  case_title = "Irradiated", # default = "Case"
  control_title = "Control", # default = "Control"
  low = "blue", # default = "green"
  mid = "#FFFFBF", # default = "black"
  high = "red" # default = "red"
)

```

In total `r nrow(scoreMatrix)` pathways were significantly altered. This high amount of regulated pathways impedes interpretaion especially due to a substantial overlap between certain pathways and biological processes. Therefore we chose to analyse only representative pathways. PathfindR offers an option to cluster pathways based on shared genes and assign a representative pathway to each cluster based on the lowest adjusted pvalue. This leads to reduction of superfluous information. Clusters were formed separately for all upregulated and all downregulated pathways, respectively.

\newpage
The top 30 activated and top 30 repressed pathways (based on a rank which was calculated as the square root of the fold-enrichment and the inverse logarithm of the pvalue) are shown in a wordcloud to facilitate biological inferences.

```{r pathway_representative_wikiPW_wordcloud, fig.width = 15, fig.height=20, fig.cap = "Wordcloud of representative pathways. Top30 activated and repressed pathways are shown."}
ggwordpath(representativeDf, z = "Direction", score = "rank", dec = TRUE, max_size = 50, n = 30, nrow = 2)
```

\newpage
As representative pathways might be less known than other pathways in each cluster we also printed the clusters with the highest number of pathways in each respective cluster. 

```{r pathway_representative_wikiPW_up, fig.width = 8.3, fig.height=11.7, fig.cap = "Activated pathways clustered by amount of shared genes."}
largeUpClusters <- as.numeric(names(table(upClustered$Cluster)[table(upClustered$Cluster)>3]))
enrichment_chart(upClustered[upClustered$Cluster %in% largeUpClusters,], plot_by_cluster = TRUE)
```

\newpage
```{r pathway_representative_wikiPW_down, fig.width = 8.3, fig.height=11.7, fig.cap = "Repressed pathways clustered by amount of shared genes."}
largeDownClusters <- as.numeric(names(table(downClustered$Cluster)[table(downClustered$Cluster)>3]))
enrichment_chart(downClustered[downClustered$Cluster %in% largeDownClusters,], plot_by_cluster = TRUE)
```


\newpage
# Appenix

## Session Info
```{r}
utils::sessionInfo()
```
