---
title: "Next-generation sequencing - analysis with DESeq2"
subtitle: "Comparison of gene expression in DSCs compared to melanocytes - differential expression analysis"
author: "Marc Bender"
date: '`r Sys.Date()`'
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../Results",
      output_format = "pdf_document"
    )
  })
output: 
  word_document:
    toc: true
    keep_md: yes
  pdf_document:
    toc: true
    keep_md: yes
  html_notebook:
    theme: united
    toc: true
  html_document:
    df_print: paged
    toc: true
    code_folding: "hide"
header-includes: 
- \usepackage[utf8]{inputenc} 
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo    = FALSE,
  dpi = 600,
  dev = c("cairo_pdf", "png")
)
```

```{r load_libraries, include = FALSE}
library(DESeq2)
library(tidyverse)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(EnhancedVolcano)
library(pheatmap)
library(ComplexHeatmap)
library(ggprism)
library(RColorBrewer)
library(circlize)
```

```{r custom_functions, include = FALSE}
marker_gene_plot <- function(gene.ids){
  markerInd   <- which(allRes$SYMBOL %in% gene.ids)
  markerRld <- assay(rld)[markerInd,]
  markerDEG <- allRes[markerInd,]
  rownames(markerRld) <- markerDEG$SYMBOL
  
  markerPlotDf <- markerRld %>%
  as.data.frame() %>% 
  rownames_to_column("SYMBOL") %>%
  gather("ID", "RLOG", -"SYMBOL") %>%
  mutate(cell_type = ifelse(str_detect(ID, "DSC"), "DSC", "Melanocytes")) %>%
  left_join(markerDEG[,c("SYMBOL", "padj")]) %>%
  mutate(padj = paste("p.adj =",formatC(padj, format = "e", digits = 2)))
  markerPlotDf[-c(1:length(gene.ids)),]$padj <- NA
  
  markerPlotDf %>%
  ggplot(aes(x= cell_type, y = RLOG, fill = cell_type)) +
  geom_bar(stat = "summary", fun = "mean", color = "black") +
  geom_text(aes(label = padj), x =1.5, y = 21, size = 3)+
  geom_point(size = 2, position = position_jitterdodge()) +
  facet_wrap(~SYMBOL, scales = "free_y") +
  ggsci::scale_fill_npg() +
  scale_x_discrete(guide = "prism_offset") +
  scale_y_continuous(limits = c(0,21), guide = "prism_offset") +
  xlab("") +
  ylab("Rlog counts") +
  guides(fill=guide_legend(title="Cell type")) +
  theme_prism(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
}
```

```{r load_data}
setwd("..")
load(file = "DSC_NGS.RData")
```

\newpage 
# Resources and introduction

This document includes the differential expression analysis for the comparison of RNASeq data from dermal stem cells (DSCs) and Melanocytes. The exploratory data analysis and functional analysis (pathway analysis and GO terms) are included in separate files. Statistical methods are described in a separate .docx file. 

For additional info on analysis workflows check: 
DESEQ workflow: https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html 
https://bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html  
https://www.bigbioinformatics.org/r-and-rnaseq-analysis 
https://github.com/bigbioinformatics/r-programming-and-rnaseq-workshop 
ClusterProfiler: https://pubmed.ncbi.nlm.nih.gov/34557778/ 

Sanity Check: 
```{r Sanity_check}
if(nrow(assay(dds)) == nrow(assay(rld)) & nrow(assay(dds)) == nrow(allRes) & nrow(assay(dds)) == nrow(res)){
  print("DESeq object, transformed count matrix and results data.frame all have the same length. Sanity check passed.")
} else {
  print("One of the objects DEseq, transformed count matrix or results data.frame is different from the others. Sanity check failed.")
}
```

\newpage
# Differential expression analysis

## Overview

```{r overview}
summary(res) %>% knitr::kable()
```

\newpage
## Volcano Plot (most differentially expressed genes labelled)

Volcano plots are commonly used to display the results of RNA-seq or other omics experiments. A volcano plot is a type of scatterplot that shows statistical significance (P value) versus magnitude of change (fold change). It enables quick visual identification of genes with large fold changes that are also statistically significant. These may be the most biologically significant genes. In a volcano plot, the most upregulated genes are towards the right, the most downregulated genes are towards the left, and the most statistically significant genes are towards the top.

This plot shows the most significantly changed genes (<= 1e-30, logFC >= |8|). 

```{r Volcano_plot_sig,  fig.width=15, fig.height= 13, fig.cap = "Volcano plot showing differentially expressed genes between Melanocytes and DSCs."}
# 5. Get the results and plot Volcano plot without filter within results function
res_volcano <- results(dds)

## annotate genes
anno <- AnnotationDbi::select(org.Hs.eg.db, rownames(res_volcano),
                              columns=c("ENSEMBL", "ENTREZID", "SYMBOL", "GENENAME"),
                              keytype="ENSEMBL")

# adding ENSEMBL gene ID as a column in significant differentially expression gene table.
dat_volcano <- cbind(ENSEMBL = rownames(res_volcano), res_volcano)
dat_volcano <- left_join(as.data.frame(dat_volcano), anno)
res_volcano <- NULL

## parameters for Volcano plot
pointSize <- 3.5
labSize   <- 5.5

# plot Volcano
# this only serves as diagnostic and representation tool. the actual fold change thresholds and pvalues change later due to the way the results function works
EnhancedVolcano(as.data.frame(dat_volcano), lab = dat_volcano$SYMBOL, selectLab = ifelse(dat_volcano$padj <= 1e-30 & abs(dat_volcano$log2FoldChange) >=8, dat_volcano$SYMBOL, NA),
                x = 'log2FoldChange', y = 'padj', gridlines.major = FALSE, gridlines.minor = FALSE,
                xlim = c(-21, 21), ylim = c(-5, 300), title = 'Melanocytes vs. DSCs', 
                pCutoff = pThres, FCcutoff = lfcThres, pointSize = pointSize, labSize = labSize) 
```

\newpage
## Volcano Plot (skin cell marker genes labelled)

This plot shows 24 selected genes from a publication assessing differences between Melanocytes (12 genes) and DSCs (12 genes) (Zabierowski et al. 2012) as well as 12 genes characteristic for Fibroblasts. 

```{r Volcano_plot_markers,  fig.width=15, fig.height= 13, fig.cap = "Volcano plot showing differentially expressed genes between Melanocytes and DSCs with labelled marker genes."}
## put marker genes at the end of the dataframe for better plotting
dat_volcano_aux <- dat_volcano[!dat_volcano$SYMBOL %in% SkinMarkerGenes, ]
dat_volcano_markers <- rbind(dat_volcano_aux, dat_volcano[dat_volcano$SYMBOL %in% SkinMarkerGenes, ])

## define colors
keyvals.colour <- ifelse(
  dat_volcano_markers$SYMBOL %in% McMarkerGenes, '#FA8072',
  ifelse(dat_volcano_markers$SYMBOL %in% DscMarkerGenes, 'royalblue',
         ifelse(dat_volcano_markers$SYMBOL %in% FibroMarkerGenes, "forestgreen", 
                ifelse(abs(dat_volcano_markers$log2FoldChange) > 0.58 & dat_volcano_markers$padj < 0.05,
                       "grey50", "grey80"))))
keyvals.colour[is.na(keyvals.colour)] <- "grey80"
names(keyvals.colour)[keyvals.colour == 'royalblue'] <- 'DSC markers'
names(keyvals.colour)[keyvals.colour == 'forestgreen'] <- 'Fibroblast markers'
names(keyvals.colour)[keyvals.colour == '#FA8072'] <- 'Melanocyte markers'
names(keyvals.colour)[keyvals.colour == 'grey50'] <- 'Differentially expressed'
names(keyvals.colour)[keyvals.colour == 'grey80'] <- 'NS'

## define alph level
keyvals.alpha <- ifelse(keyvals.colour %in% c("royalblue", "#FA8072", "forestgreen"), 1, 0.4)

EnhancedVolcano(as.data.frame(dat_volcano_markers), lab = dat_volcano_markers$SYMBOL, selectLab = SkinMarkerGenes,
                x = 'log2FoldChange', y = 'padj', gridlines.major = FALSE, gridlines.minor = FALSE,
                xlim = c(-21, 21), ylim = c(-5, 300), title = 'Melanocytes vs. DSCs', colCustom = keyvals.colour,
                colAlpha = keyvals.alpha, drawConnectors = TRUE, widthConnectors = 1.0, colConnectors = 'black', 
                arrowheads = FALSE, max.overlaps=Inf,
                pCutoff = pThres, FCcutoff = lfcThres, pointSize = pointSize, labSize = labSize)
```

 \newpage
## Volcano Plot (melanoma marker genes labelled)

This plot shows genes frequently mutated in melanoma [taken from Cherepakhin et. al (2022)] and differentially expressed genes [taken form Xie et. al (2022)]. In total `r nrow(MelanomaMarkerGenes)` differentially regulated (or mutationally amplified/repressed) genes were either labelled as "upregulated in melanoma" or "downregulated in melanoma". Out of these `r nrow(MelanomaMarkerGenes)` genes, `r sum(allRes$SYMBOL %in% MelanomaMarkerGenes$SYMBOL)` genes were expressed above the expression threshold. The following table shows the total number of genes which were identified to be up/downregulated in melanoma and the corresponding results of up/downregulation in our experiments: 

`r knitr::kable(crossTab)`

\newpage
```{r Volcano_plot_melanoma,  fig.width=15, fig.height= 13, fig.cap = "Volcano plot showing differentially expressed genes between Melanocytes and DSCs with tumour marker genes."}
##
dat_volcano_aux <- dat_volcano[!dat_volcano$SYMBOL %in% MelanomaMarkerGenes$SYMBOL, ]
dat_volcano_markers <- rbind(dat_volcano_aux, dat_volcano[dat_volcano$SYMBOL %in% MelanomaMarkerGenes$SYMBOL, ])

## define colors
keyvals.colour <- ifelse(dat_volcano_markers$SYMBOL %in%  MelanomaMarkerGenes[MelanomaMarkerGenes$TYPE == "Melanoma_up",]$SYMBOL, '#FA8072',
         ifelse(dat_volcano_markers$SYMBOL %in% MelanomaMarkerGenes[MelanomaMarkerGenes$TYPE == "Melanoma_down",]$SYMBOL, "royalblue", 
                ifelse(abs(dat_volcano_markers$log2FoldChange) > 0.58 & dat_volcano_markers$padj < 0.05,
                       "grey50", "grey80")))
keyvals.colour[is.na(keyvals.colour)] <- "grey80"
names(keyvals.colour)[keyvals.colour == '#FA8072'] <- 'Melanoma up'
names(keyvals.colour)[keyvals.colour == 'royalblue'] <- 'Melanoma down'
names(keyvals.colour)[keyvals.colour == 'grey50'] <- 'Differentially expressed'
names(keyvals.colour)[keyvals.colour == 'grey80'] <- 'NS'

## define alph level
keyvals.alpha <- ifelse(keyvals.colour %in% c("black", "forestgreen"), 1, 0.4)

EnhancedVolcano(as.data.frame(dat_volcano_markers), lab = dat_volcano_markers$SYMBOL, selectLab = MelanomaMarkerGenes$SYMBOL,
                x = 'log2FoldChange', y = 'padj', gridlines.major = FALSE, gridlines.minor = FALSE,
                xlim = c(-21, 21), ylim = c(-5, 300), title = 'Melanocytes vs. DSCs', colCustom = keyvals.colour,
                colAlpha = keyvals.alpha, drawConnectors = TRUE, widthConnectors = 1.0, colConnectors = 'black', 
                arrowheads = FALSE, max.overlaps=10,
                pCutoff = pThres, FCcutoff = lfcThres, pointSize = pointSize, labSize = labSize)
```

\newpage
## MA plot

An MA-plot (Dudoit et al. 2002) provides a useful overview for the distribution of the estimated
coefficients in the model, e.g. the comparisons of interest, across all genes. On the y-axis, the “M”
stands for “minus” – subtraction of log values is equivalent to the log of the ratio – and on the x-axis,
the “A” stands for “average”. You may hear this plot also referred to as a mean-difference plot, or a
Bland-Altman plot.

Before making the MA-plot, we use the lfcShrink function to shrink the log2 fold changes for the comparison
of DSCs vs. Melanocytes. There are three types of shrinkage estimators in DESeq2, which are covered
in the DESeq2 vignette. Here we specify the apeglm method for shrinking coefficients, which is good for shrinking
the noisy LFC estimates while giving low bias LFC estimates for true large differences (Zhu, Ibrahim, and Love 2018).
To use apeglm we specify a coefficient from the model to shrink, either by name or number as the coefficient appears
in resultsNames(dds).

```{r MA_plot, fig.width=5, fig.height= 4, fig.cap = "MA plot with original log fold changes."}
### plot MA plot

# plot with original lfc
plotMA(res, ylim = c(-10, 10), alpha = pThres, colSig = "red2", colNonSig = "grey30", main = "MA plot with original log fold changes")
```

<br> 

```{r MA_plot_shrunk, fig.width=5, fig.height= 4, fig.cap = "MA plot with shrunken log fold changes."}
# shrink estimates and exlpore differences in MA plots
plotMA(resShrunk, ylim = c(-10, 10), alpha = pThres, colSig = "red2", colNonSig = "grey30", main = "MA plot with shrunken log fold changes")

###
# WARNING: DO NOT ALWAS SHRINK EVERYTHING BY DEFAULT. Look at MA plot and Volcano plot to see if it makes sense
### 
```

\newpage
## Comparison of marker genes

## Expression of melanocyte markers

Marker genes were labelled according to the publication of Zabierowski 2012, lab practice and https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4278762/. 

```{r mc_marker_genes, fig.width=7, fig.height= 8, fig.cap = "Expression of melanocyte markers."}
mcMarkers <- c("MLANA", "PMEL", "TYR", "TYRP1", "MITF", "DCT",  "KIT", "CDH1", "ERBB3", "EDNRB", "SOX10", "PAX3")
marker_gene_plot(mcMarkers)
```

\newpage
## Expression of dsc markers

```{r dsc_marker_genes, fig.width=7, fig.height= 8, fig.cap = "Expression of DSC markers."}
dscMarkers <-  c("SNAI1", "NGFR", "NES", "FRZB", "FOXC1", "JAG1", "PAX3", "HES4", "MSX1", "S100A4", "SOX2", "CDH2")
marker_gene_plot(dscMarkers)
```

\newpage
## Expression of fibroblast markers

```{r fb_marker_genes, fig.width=7, fig.height= 8, fig.cap = "Expression of fibroblast markers."}
fbMarkers <- c("THY1", "DLK1", "FAP", "PDGFRA", "ELN", "FN1", "VIM", "LAMA5", "ITGB1", "DPP4", "SERPINH1", "LAMA2")
marker_gene_plot(fbMarkers)
```



\newpage
## Heatmap

If using either of Euclidean distance or Pearson correlation, data should follow a Gaussian /
normal (parametric) distribution. So, if coming from a microarray, anything from RMA normalisation is fine,
whereas, if coming from RNA-seq, any data deriving from a transformed normalised count metric should be fine,
such as variance-stabilised, regularised log, or log CPM expression levels.

If you are performing clustering on non-normal data, like 'normalised' [non-transformed] RNA-seq counts,
FPKM expression units, etc., then use Spearman correlation (non-parametric).

```{r Heatmap_function}
## Wrapper around Heatmap function for quick generation of ComplexHeatmaps
drawHeatmap <- function(matScaled, title = paste("Differentially Expressed genes (", nrow(sigRes), ")", sep =""), 
                        show_row_names = FALSE, show_column_names = TRUE,...){
  Ht <- Heatmap(matScaled, col = col_fun,  top_annotation = annColors, column_title = title,
                column_title_gp = gpar(fontsize = heatmapFontsize, fontface = "bold"),  
                show_column_names = show_column_names,
                show_row_names = show_row_names, 
                heatmap_legend_param = list(title_gp = gpar(fontsize = heatmapFontsize),                                                                     labels_gp = gpar(fontsize = heatmapFontsize),
                title = "Row Z-score", at = seq(-2,2,by=1),  color_bar="continuous",title_position ="topleft", 
                legend_height = unit(4, "cm")), ...)
  draw(Ht, merge_legend = TRUE)
}
```

\newpage
### All differentially expressed (DE) genes
```{r Heatmap, fig.width=12, fig.height= 14, fig.cap = "Heatmap of all differentially expressed genes."}
## specify fontsize
heatmapFontsize <- 18

# Heatmap of all DE genes
matAll <- assay(rld)[sigRes$ENSEMBL,]
## scale matrix
datScaled <- matAll %>%  t() %>% scale() %>%  t()

## specify annotation bar
annoHeatmap <- as.data.frame(colData(rld)[, c("cell_type")])
colnames(annoHeatmap) <- "Cell_type"

## specify annotation bar colors
annColors <- HeatmapAnnotation(df =annoHeatmap,
                               col = list(Cell_type = c("DSC" = ggsci::pal_npg("nrc")(2)[1], "Melanocytes" =
                                                          ggsci::pal_npg("nrc")(2)[2])),
                               annotation_legend_param = list(Cell_type = list(nrow=1), 
                                                              title_gp = gpar(fontsize = heatmapFontsize), 
                                                              labels_gp = gpar(fontsize = heatmapFontsize)),
                               annotation_name_gp = gpar(fontsize = heatmapFontsize, fontface = "bold"))

## define color function
col_fun <-  colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(100)

## print heatmap
drawHeatmap(datScaled, title = paste("Differentially Expressed genes (", nrow(sigRes), ")", sep =""))
```

\newpage
### Top 30 DE genes
```{r HeatmapTop30, fig.width=12, fig.height= 11, fig.cap = "Heatmap of the Top30 differentially expressed genes."}
nGenes <- 30
topSignif <- head(order(allRes$padj, decreasing = FALSE),nGenes)
labelsTop30 <- allRes[topSignif,]$SYMBOL
matTop30  <- assay(rld)[topSignif, ]
rownames(matTop30) <- labelsTop30
datScaled <- matTop30 %>%  t() %>% scale() %>%  t()

drawHeatmap(datScaled, title = paste("Top ",nGenes," differentially expressed genes", sep =""), show_row_names = TRUE)
```


\newpage
### Comparison of genes identified with the DRAGEN pipeline

This plot includes the same 30 genes which are depicted in the Heatmap included in the output generated from the DRAGEN analysis software. 

```{r HeatmapDragen, fig.width=12, fig.height= 11, fig.cap = "Heatmap of genes identified as Top30 differentially expressed in the DRAGEN analysis pipeline."}
ind <- which(allRes$SYMBOL %in% DragenRes$gene_name)
labelsDragen <- allRes[ind,]$SYMBOL
matDragen  <- assay(rld)[ind, ]
rownames(matDragen) <- labelsDragen
datScaled <- matDragen %>%  t() %>% scale() %>%  t()

drawHeatmap(datScaled, title = paste("Top ",nGenes," differentially expressed genes", sep =""), show_row_names = TRUE)
```

\newpage
# Appenix

## Session Info
```{r}
utils::sessionInfo()
```
